---
title: "Introduction to Crash Data Wrangling, Analyses, and Visualization"
author: "Presenter - Josh Roll"
date: "Workshop Date - September 14th, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, echo = FALSE, results = "hide"}
##Load libraries
	library(httr)
	suppressWarnings(library(knitr))
	library(jpeg)
  library(png)
  library(grid)
```


###Fatal Traffic Injuries in Oregon
####How is Oregon doing in reducing traffic fatal injuries?

####Based on the counts of fatal injuries from ODOT data for 2002 to 2016, fatal injuries in Oregon appear to be near historic highs.  But what would a longer view look like?


```{r, echo = FALSE,fig.cap=paste("Source: ODOT Crash Data (2002-2016)"),fig.width=16, fig.height=8}
img1_path <- "Z:/JRoll/Workshops/TC Data Science Workshop 2018/Crash Exercise/Graphics/Oregon_Traffic_Fatals_2002_2016.jpeg"
img1 <- readJPEG(img1_path)
#attr(img1, "info")
grid.raster(img1)
```


<br>


####Oregon DOT doesn't have readily available data before 2002, so what's a traffic safety analyst to do?

####Luckily, the National Highway Transprotation Safety Administration(NHTSA) maintains the Fatal Accident Reporting System (FARS) and the data have been collected since 1975.


```{r, echo = FALSE,fig.cap=paste("Source: NHTSA - Fatal Accident Reporting System (1975-2016)"),fig.width=16, fig.height=8}
img1_path <- "Z:/JRoll/Workshops/TC Data Science Workshop 2018/Crash Exercise/Graphics/Oregon_Traffic_Fatals_1975_2016.jpeg"
img1 <- readJPEG(img1_path)
#attr(img1, "info")
grid.raster(img1)
#include_graphics(img1_path)
```

<br>

####Viewing these data with a longer term perspective shows the recent increase is still lower than historic highs in the late 1970s but are now about where we were in the 1990s.  Now lets investigate how we created these graphics and probe these data further.  

<br>













